Practical Optional Types for Clojure

### 2015-02-27 overview.tex
move syntax descriptions to overview

\subsection{Clojure and Typed Clojure Primer}

We now review the basics of the Clojure subset we use in the paper.

\paragraph{Syntax} Parentheses indicate \emph{applications}, brackets
delimit
\emph{vectors}, braces
delimit
\emph{hash-maps}
and double quotes delimit Java \emph{strings}.
\clj{nil} is exactly Java's \java{null}.
\emph{Symbols} begin with an alphabetic character.
A colon prefixed symbol like \clj{:a} is a \emph{keyword}.
\emph{Commas} are always whitespace.

\paragraph{Core forms} \clj{if} is a standard two-armed conditional.
\clj{do} sequences two expressions returning the latter.
\clj{throw} corresponds to Java's \java{throw}.
\clj{fn} creates a function from a parameter vector and a body expression.
\clj{let} takes a flat binding vector (which can refer to previous bindings like Scheme's \mintinline{scheme}{let*})
and a body expression.
\clj{def} binds a top-level \emph{var}.
\clj{defn} combines \clj{def} and \clj{fn}.

\paragraph{Semantics} \clj{get} applied to a hash-map and a key returns the value
associated with the key in the map, or \clj{nil} if there is none.
Keywords are functions that apply \clj{get} to their argument
and itself.
\clj{inc} and \clj{dec} increment and decrement numbers;
\clj{assoc} and \clj{dissoc} associate and dissociate entries
from immutable maps.
\clj{class} returns the argument's class, or \clj{nil} if passed \clj{nil}.
The \clj{=} function returns true if both arguments are equal.
\clj{number?} is
a Java \java{instanceof} test of \clj{java.lang.Number}.
\clj{nil} and \clj{false} are the false values.

\paragraph{Namespaces} \clj{ns}
declares a
\emph{namespace}
which manage var and class imports. 
\clj{:require} takes a number of namespace dependencies and \clj{:refer}
brings vars into scope by name.
\clj{:refer-clojure} \clj{:exclude} unmaps core vars from the current namespace.

\paragraph{Java Interoperability} Java methods and fields have prefix notation
like \clj{(.method target args*)} and \clj{(.field target)} respectively,
with method and field names prefixed with a dot and methods taking some number of arguments.
Constructors are called with the class name with a dot suffix and arguments like \clj{(class. args*)}.

\paragraph{Metadata} Symbols support \emph{metadata}---the syntax \clj{^File x}
is a single expression of a symbol \clj{x} with metadata \clj{{:tag File}}.
In binding positions like \clj{(fn [^File x] ...)}
syntactic occurrences preserve metadata.

\paragraph{Destructuring} \emph{Map destructuring} is binding position syntax for pattern matching.
Parameters are replaced with maps of symbols to keywords that bind the symbols
to lookups on that keyword, optionally terminated by \clj{:as}
which aliases the parameter. For example
\clj{(fn [{^File x :x, :as m}] ...)}
expands to a \clj{let} binding \clj{m} to the first argument and 
\clj{^File x} to \clj{(:x m)}.

\paragraph{Static type syntax} Single quoted keywords are singleton types containing just the keyword.
Single quoted maps mapping keywords to types 
are HMap types representing
maps with known keyword entries with values of the associated type.
\clj{U} creates an \emph{untagged union} type, which can take
any number of types.

\paragraph{Typed Clojure Macros}
The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
The typed \clj{fn} supports optional annotations by 
adding
\clj{:-} and a type after a parameter
position
or binding vector 
to annotate parameter types
and return types respectively.

@@

Here is above program in Typed Clojure.

@@

This is a regular Clojure file compiled with
the Clojure compiler. It has a
runtime dependency 
on
\clj{clojure.core.typed},
Typed Clojure's core namespace.
Programmers import typed versions of `typed' macros and qualified types to ease annotation.
The \clj{check-ns} function type checks the current namespace.

Typed Clojure can check all the examples in~\citet{TF10}---the 
rest of this section describes the extensions necessary
to check Clojure code.

@@

In this example a \clj{throw} expression guards \clj{(inc x)}
from being evaluated if \clj{x} is \clj{nil}, preventing
a possible null-pointer exception.

@@

Here \clj{defalias} defines a type abbreviation called \clj{Expr}
that describes the structure of a recursively-defined AST as a union of HMaps.
A quoted keyword in a type such as \clj{':if} is a singleton type that contains just the keyword.
\clj{an-exp} is a function verified to return an \clj{Expr}.

@@

\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
dissociating the raw \clj{:private-key} entry and associating an encrypted private key
as \clj{:enc-private-key}.
Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
operation forgot to remove it.

The next example is the same except we use the \clj{:absent-keys}
HMap option
and
utilize destructuring.

@@

We would still catch the bug if \clj{dissoc} forgot to remove \clj{:private-key},
but now stashing the raw private key as \clj{:secret-key} is allowed.

\paragraph{Branching on HMaps} Finally, testing on HMap properties
allows us to refine its type down branches. \clj{dec-map} takes an
\clj{Expr}, traverses to its nodes and decrements their values, then
builds the \clj{Expr} back up with the decremented nodes.

@@

@@

@@

A multimethod in Clojure is a function with a \emph{dispatch
function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
\inputminted[firstline=5,lastline=6]{clojure}{code/demo/src/demo/rep.clj}
The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}.

>>>>

\subsection{Clojure Syntax}

We describe new syntax as they appear in each example, but
we also include the essential basics of Clojure syntax.

\clj{nil} is exactly Java's \java{null}.
Parentheses indicate \emph{applications}, brackets
delimit
\emph{vectors}, braces
delimit
\emph{hash-maps}
and double quotes delimit \emph{Java strings}.
\emph{Symbols} begin with an alphabetic character,
and a colon prefixed symbol like \clj{:a} is a \emph{keyword}.

\emph{Commas} are always \emph{whitespace}.

@@

Here is the above program in Typed Clojure.

@@

This is a regular Clojure file compiled with
the Clojure compiler, with the \clj{ns} form
declaring a
\emph{namespace}
for managing var and class imports.
Here \clj{:require} declares a runtime dependency 
on
\clj{clojure.core.typed},
Typed Clojure's core namespace, and
\clj{:refer}
brings a collection of vars into scope by name.
The \clj{:refer-clojure} \clj{:exclude} option unmaps core vars from the current 
namespace---here we unmap \clj{clojure.core/fn},
which creates a function from a parameter vector and a body expression,
and import a typed variant of \clj{fn}
that supports type annotations.

The typed \clj{fn} supports optional annotations by 
adding
\clj{:-} and a type after a parameter
position
or binding vector 
to annotate parameter types
and return types respectively.
Typed Clojure provides a \clj{check-ns} function to type checks the current namespace.
\clj{number?} is
a Java \java{instanceof} test of \clj{java.lang.Number}.
As in Typed Racket, \clj{U} creates an \emph{untagged union} type, which can take
any number of types.

Typed Clojure can already check all of the examples in~\citet{TF10}---the 
rest of this section describes the extensions necessary
to check Clojure code.

@@

The \clj{do} form sequences two expressions returning the latter,
\clj{throw} corresponds to Java's \java{throw}
and \clj{(class. args*)} is the syntax for Java constructors---that 
is a class name with a dot suffix as the operator followed the arguments to the constructor.

In this example a \clj{throw} expression guards \clj{(inc x)}, the increment function for numbers,
from being evaluated if \clj{x} is \clj{nil}, preventing
a possible null-pointer exception.

@@

The \clj{defn} macro defines a top-level function, with syntax like the typed \clj{fn}.
The function \clj{an-exp} is verified to return an \clj{Expr}.

The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
Here \clj{defalias} defines \clj{Expr}
that describes the structure of a recursively-defined AST as a union of HMaps.
A quoted keyword in a type, such as \clj{':if}, is a singleton type that contains just the keyword.
A type that is a quoted map like \clj{'{:op ':if}} is a
HMap type with a fixed number of keyword entries of the specified types
known to be \emph{present},
zero entries known to absolutely be \emph{absent},
and an infinite number of \emph{unknown} entries entries.
Since only keyword keys are allowed, they do not require quoting.

@@

\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
non-destructively dissociating the raw \clj{:private-key} entry with \clj{dissoc}
and associating an encrypted private key
as \clj{:enc-private-key} on an immutable map with \clj{assoc}.
The expression \clj{(:private-key kp)} shows that keywords are also 
functions that look themselves up in a map returning the associated value or \nil{} if the key is missing.
Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
operation forgot to remove it.

The next example is the same except we use the \clj{:absent-keys} HMap option.
We also
utilize \emph{map destructuring}, which is binding position syntax for pattern matching.
Parameters are replaced with maps of symbols to keywords that bind the symbols
to lookups on that keyword, optionally terminated by \clj{:as}
which aliases the parameter. For example
\clj{(fn [{^File x :x, :as m}] ...)}
expands to a \clj{let} binding \clj{m} to the first argument and 
\clj{^File x} to \clj{(:x m)}.
Clojure's \clj{let} takes a flat binding vector which can refer to previous bindings like Scheme's \mintinline{scheme}{let*},
and a body expression.

@@

Since this example enforces that \clj{:private-key} must not appear
in a \clj{EncKeyPair}
Typed Clojure would still complain if we forgot to \clj{dissoc} \clj{:private-key}
from the return value.
Now, however we could stash the raw private key in another entry
like \clj{:secret-key} which is not mentioned by the partial HMap \clj{EncKeyPair}
without Typed Clojure noticing.

\paragraph{Branching on HMaps} Finally, testing on HMap properties
allows us to refine its type down branches. \clj{dec-map} takes an
\clj{Expr}, traverses to its nodes and decrements their values by \clj{dec}, then
builds the \clj{Expr} back up with the decremented nodes.

@@

Calls to Java methods and fields have prefix notation
like \clj{(.method target args*)} and \clj{(.field target)} respectively,
with method and field names prefixed with a dot and methods taking some number of arguments.

@@

Symbols support \emph{metadata}---the syntax \clj{^File f}
is a single expression that is a symbol \clj{f} with metadata \clj{{:tag File}}.
In binding positions like \clj{(fn [^File f] ...)}
syntactic occurrences preserve metadata.

@@

A multimethod in Clojure is a function with a \emph{dispatch
function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
\inputminted[firstline=5,lastline=6]{clojure}{code/demo/src/demo/rep.clj}
The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}, a function that
returns the class of its argument or \clj{nil} if passed \clj{nil}.
